---
title: "Creating a global land mask raster"
author: "Rich Cottrell"
date: "25/10/2022"
output:
  pdf_document: default
  html_document: default
---

Libraries
```{r, eval=TRUE, results='hide', message=FALSE,echo=TRUE}
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(here)
library(raster)

```


First get the shapefile for the land from Natural Earth
```{r}

land_sf <- ne_countries(scale = "medium", returnclass = "sf")


```

Now get a blank raster in the projection and resolution you want (using the default wgs84 here but in 1 degree). The output raster can then be multiplied with your raster of interest to return all non-land cells. If you want to keep only land cells, replace inverse =TRUE with inverse = FALSE in the mask call. 

```{r}
base_r <- rast(res = 1)
terra::values(base_r) <- 1

#this creates a raster with na values for the land (1 elsewhere)
land_mask_terra <- terra::mask(base_r, vect(land_sf), inverse=TRUE)

#retains 1 for land
plot(land_mask_terra)

writeRaster(x = land_mask_terra, filename = here("data/landmask_spatrast.tif"), overwrite = TRUE)

```

This time in raster format
```{r}

base_r <- raster(res = 1)
raster::values(base_r) <- 1

#this creates a raster with na values for the land (1 elsewhere)
land_mask_raster <- raster::mask(base_r, land_sf, inverse=TRUE)

plot(land_mask_raster)

raster::writeRaster(x = land_mask_raster, filename = here("data/landmask_rasterlayer.tif"), overwrite = TRUE)


```

